Distributed Algorithms for Sparse Matrix-Vector Multiplication

### 2015-04-22 presentation.Rmd
First paper complete... ish

$Y = X\beta + \epsilon$

@@

- Y is an $n \times 1$ response vector
- X is an $n \times k$ design matrix
- $\beta$ is a $k \times 1$ parameter vector
- $\epsilon$ is an $n \times 1$ error vector (that is, $\epsilon \sim MVN(0, \Sigma)$)

@@

$Y = \begin{bmatrix}
        18 & 21 & 20 & 21 & 16
     \end{bmatrix}^{T}$

@@

>>>>

$y = X\beta + \epsilon$

@@

- y is an $n \times 1$ response vector
- X is an $n \times k$ design matrix
- $\beta$ is a $k \times 1$ parameter vector
- $\epsilon$ is an $n \times 1$ error vector (that is, $\epsilon \sim MVN(0, \Sigma)$)

## Least Squares Estimator
The "Least Squares Estimator" of the parameter vector $\beta$ is given by $b = (X^TX)^{-1}X^Ty$

@@

$y = \begin{bmatrix}
        18 & 21 & 20 & 21 & 16
     \end{bmatrix}^{T}$

@@

## A Simple Example (Continued)
$(X^TX)^{-1} = \begin{bmatrix}
        60.2714 & -2.0714 \\
        -2.0714 & 0.0714 \\
     \end{bmatrix}$

$X^Ty = \begin{bmatrix}
        96 \\
        2797 \\
     \end{bmatrix}$

$b = (X^TX)^{-1}X^Ty = \begin{bmatrix}
        -7.7286 \\
        0.9286 \\
     \end{bmatrix}$

Hence, the equation for the line of best fit is given by $y = -7.7286 + 0.9286x$.

## SpMV Calculations
Note one of the key operations here is multiplying $X^T$ by $y$. In this example, it was almost trivial, but in general, the matrix $X$ can be very large and sparse.

In particular, there is a column in $X$ for every parameter in the parameter vector.

## R Simulation
```{r, eval=FALSE}
wearables <- read.csv("data/Example_WearableComputing_weight_lifting_exercises_biceps_curl_variations.csv")
X <- apply(wearables[,-1], 2, as.numeric)
y <- as.numeric(wearables[,1])

biggerX <- do.call("rbind", replicate(50, X, simplify = FALSE))
biggery <- rep(y, times = 50)

t(X) %*% y
t(biggerX) %*% biggery
```

## Limitations of the R (and naive) implementation
- No explicit parallelism (only thread level) by default
- Even if it was explicitly parallel, no load balancing or processor communication
- Choice of multi-core or multi-machine parallelism
- Poor cache locality

## Load Balanced Sparse Matrix-Multiplication
Nastea, Frieder, and El Ghazawi (1997) published one of the earliest proposed solutions to this issue.

## LBSM-M Overview
- $Y_i = AX_i$ where A is the sparse matrix, and $X_i$ is a sequence of dense vectors
- The size of the sequence of $X_i$ vectors is very large and not a priori known
- The resulting $Y_i$ vectors are generated and transmitted on an individual basis

## Load Balancing 
Greedy allocation for sparse pattern matrices:

$F = max_i\{\sum_{j=1}^M(nZ_{i_j})\}$

$i = 1, 2, ..., P$ represents the processors.

$i_j = \{i_1, i_2, ..., i_M\}$ represents indices of rows assigned to processor i

$nZ_{i_j}$ is the number of non-zero elements in these rows.

The goal is to minimize the function F, which is minimizing the **largest bucket size that yields the highest computing time**.

## Highly Skewed Data
In a matrix that is both highly skewed and highly sparse, splitting rows that have a significantly larger number of non-zero elements is necessary. This incurs a slight overhead and hence is only recommended for more extreme cases.

## Experimental Results
Synthetically generated data:

$f(i) = \frac{C}{i^{1 - \theta}} \text{ for } i = \{1, 2, ..., N\}$

$C = \frac{1}{\sum_{i=1}^N\frac{1}{i^{1 - \theta}}}$

$\theta \in [0, 1]$

This is a discrete distribution called the **Zipf Distribution**

## Experimental Results (Continued)
<img src="images/im1.png" width="800px">

## Experimental Results (Continued)
<img src="images/im2.png" width="800px" height="500px">

## Conclusion

## Second Paper
An architecture-aware technique for optimizing sparse matrix-vector multiplication on GPUs by Maggioni and Berger-Wolf

## Sparse Matrix Compression
<img src="images/im3.png" width="800px">

## Cache Heuristic

## Cache Heuristic (Continued)

## Results

## Results (Continued)

## Conclusion

## Third Paper

## The BRC (Blocked Row-Column) Format

## Model Parameters

## Results

## Results (Continued)

## Multiple GPUs

## Caveat: Pre-Processing Overhead

## Conclusion
